---
title: "main"
author: "Nasim Bondar Sahebi"
date: "2023-11-07"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## load libraries

```{r libraries}
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(dplyr)
library(knitr)
library(kableExtra)
library(tools)
```


## Data Loading and Preparation

This section lists and displays all `.rda` files from the `data` and `data_sex` directories, representing 47 study cohorts.


```{r  all data}

# Set base directory
Base_dir <- "~/BHK lab/Ecotyper/"

# List .rda files in both directories
data_files <- list.files(path = file.path(Base_dir, "data"), pattern = "\\.rda$", full.names = TRUE)
data_sex_files <- list.files(path = file.path(Base_dir, "data_sex"), pattern = "\\.rda$", full.names = TRUE)

# Display files
data_files_df <- data.frame(Files = basename(data_files)) #47 study
kable(data_files_df, caption = "All studies", col.names = "Cohorts")

```

## Combine all Datasets

This section loads and combines clinical datasets from the `data` and `data_sex` directories. Key steps include:

- Listing `.rda` files from both directories.
- Extracting and merging clinical data based on common columns.
- Creating unique study and patient identifiers.
- Displaying combined clinical data in tables.

The final datasets are saved as `.rds` files for future use.


```{r Combine clinicaldata 1 }


# Set base directory
Base_dir <- "~/BHK lab/Ecotyper/"

# List .rda files in both directories
data_files <- list.files(path = file.path(Base_dir, "data"), pattern = "\\.rda$", full.names = TRUE)
data_sex_files <- list.files(path = file.path(Base_dir, "data_sex"), pattern = "\\.rda$", full.names = TRUE)

# Display data files
data_files_df <- data.frame(Files = basename(data_files)) # List all study .rda files
kable(data_files_df, caption = "All studies", col.names = "Cohorts") 

# Initialize lists for storing column names and clin dataframes for data_files
column_list <- list()
clin_list <- list()

for (data_path in data_files) {
    obj <- load(data_path)  
    se <- get(obj)
    
    clin <- as.data.frame(colData(se))  
    
    # Set row names with base name and index
    base_name <- tools::file_path_sans_ext(basename(data_path))
    rownames(clin) <- paste0(base_name, "_", seq_len(nrow(clin)))
    
    column_list[[data_path]] <- colnames(clin)  # Store column names
    clin_list[[data_path]] <- clin  # Store clinical data
}

# Identify common columns across all datasets
common_columns <- Reduce(intersect, column_list)

# Bind all clin dataframes using the common columns
combined_clin <- do.call(rbind, lapply(clin_list, function(clin) {
    clin[, common_columns, drop = FALSE]  # Subset to common columns
}))

# Create study and patient identifiers
combined_clin$study <- gsub("\\..*","", basename(rownames(combined_clin)))
combined_clin$study_patient <- paste0(combined_clin$study, "_", combined_clin$patientid)
rownames(combined_clin) <- combined_clin$study_patient

# Repeat for data_sex_files
data_sex_files_df <- data.frame(Files = basename(data_sex_files)) # List all data_sex .rda files

# Initialize lists for storing column names and clin dataframes for data_sex
column_sex_list <- list()
clin_sex_list <- list()

for (data_sex_path in data_sex_files) {
    obj <- load(data_sex_path)  
    clin_sex <- get(obj)
    
    # Set row names with base name and index
    base_name <- tools::file_path_sans_ext(basename(data_sex_path))
    rownames(clin_sex) <- paste0(base_name, "_", seq_len(nrow(clin_sex)))
    
    column_sex_list[[data_sex_path]] <- colnames(clin_sex)  # Store column names
    clin_sex_list[[data_sex_path]] <- clin_sex  # Store clinical data
}

# Identify common columns across all data_sex datasets
common_sex_columns <- Reduce(intersect, column_sex_list)

# Bind all clin_sex dataframes using the common columns
combined_clin_sex <- do.call(rbind, lapply(clin_sex_list, function(clin_sex) {
    clin_sex[, common_sex_columns, drop = FALSE]  # Subset to common columns
}))

# Create study and patient identifiers for data_sex
combined_clin_sex <- combined_clin_sex[!is.na(combined_clin_sex$patientid), ]
combined_clin_sex$study <- gsub("\\..*","", basename(rownames(combined_clin_sex)))
combined_clin_sex$study_patient <- paste0(combined_clin_sex$study, "_", combined_clin_sex$patientid)
rownames(combined_clin_sex) <- combined_clin_sex$study_patient


kable(head(combined_clin), caption = "Combined Clinical Data from Both Datasets", format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

# Display the combined_clin table
kable(head(combined_clin),
      caption = "Summary of Combined Clinical Data (Data Directory)",
      col.names = colnames(combined_clin),
      format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

# Display the combined_clin_sex table
kable(head(combined_clin_sex),
      caption = "Summary of Combined Clinical Data (Data_Sex Directory)",
      col.names = colnames(combined_clin_sex),
      format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)


# saveRDS(combined_clin, file = file.path("~/BHK LAB/Ecotyper/outputs/", "combined_clin.rds"))

# saveRDS(combined_clin_sex, file = file.path("~/BHK LAB/Ecotyper/outputs/", "combined_clin_sex.rds"))


```

## Generate Table

This section of the analysis generates a summary table of clinical characteristics from the combined clinical dataset. The table includes the number of studies, patients, age (with median and IQR), missing values (NA %), sex distribution, cancer types, treatments, response rates, and survival outcomes (PFS and OS). The function `generate_all_data()` processes the data and computes these metrics for both the full dataset and the dataset with sex-specific information.

```{r Table 1 }

Characteristic <- c("N° studies (%)", "N° patients", "Age year", "Median (IQR)", "NA (%)", "Sex", "Female (%)", "Male (%)", "NA (%)", "Cancer Type","Kidney (%)", "Brain (%)","Bladder (%)", "Lung (%)", "Melanoma (%)", "Liver (%)", "Lymph_node (%)", "Unknown (%)","Ureteral (%)", "Pancreas (%)", "HNC (%)","Colon (%)", "Breast (%)" , "Treatment", "PD-1/PD-L1 (%)", "IO+combo (%)", "CTLA4 (%)" , "IO+chemo (%)", "IO+targeted (%)", "Response", "No-response (%)", "Response (%)", "NA (%)", "PFS", "No-progression (%)", "Progressed (%)", "NA (%)", "Median PFS", "IQR PFS", "OS", "Alive (%)", "Dead (%)", "NA (%)", "Median OS", "IQR OS")


# Find the intersection of treatments and cancer types between the two datasets
treatments <- intersect(unique(combined_clin_sex$treatment), unique(combined_clin$treatment))
cancer_types <- intersect(unique(combined_clin_sex$cancer_type), unique(combined_clin$cancer_type))

# Generate data function with only the common treatments and cancer types
generate_all_data <- function(data) {
  All_DATA <- c(
    paste(length(unique(data$study)), "%"),    # N° studies          
    length(data$patientid),                    # N° patients          
    "",                                       # Age year placeholder
    
    # 1. Age year placeholder
    paste(median(data$age, na.rm = TRUE), " (", min(data$age, na.rm = TRUE), "-", max(data$age, na.rm = TRUE), ")", sep = ""),
    paste(sum(is.na(data$age)), " (", round(sum(is.na(data$age)) / nrow(data) * 100, 1), "%)", sep = ""),
        
    "",                                     
    
    # 2. Sex placeholder
    paste(sum(data$sex == "F", na.rm = TRUE) , " (",round(sum(data$sex == "F", na.rm = TRUE) / nrow(data) * 100, 1), "%)"),  # Female percentage
    paste( sum(data$sex == "M", na.rm = TRUE), " (",round(sum(data$sex == "M", na.rm = TRUE) / nrow(data) * 100, 1), "%)"),  # Male percentage
    paste(sum(is.na(data$sex)) , " (",round(sum(is.na(data$sex)) / nrow(data) * 100, 1), "%)"),  # NA percentage for sex
    "",                                       # Cancer Type placeholder
    
    # 3. Calculate common cancer type percentages
    sapply(cancer_types, function(ct) {
      paste(sum(data$cancer_type == ct), " (", round(sum(data$cancer_type == ct) / nrow(data) * 100, 1), "%)", sep = "")
    }), 
    "",                                       
    
    # 4. Treatment placeholder for common treatments
    sapply(treatments, function(tr) {
      paste(sum(data$treatment == tr), " (", round(sum(data$treatment == tr) / nrow(data) * 100, 1), "%)", sep = "")
    }), 
    "",                                      
    
    # 5. Response placeholder
    paste(sum(data$response == "NR", na.rm = TRUE), " (", round(sum(data$response == "NR", na.rm = TRUE) / nrow(data) * 100, 1), "%)"), 
    paste(sum(data$response == "R", na.rm = TRUE), " (",round(sum(data$response == "R", na.rm = TRUE) / nrow(data) * 100, 1), "%)"),
    paste(sum(is.na(data$response)), "(", round(sum(is.na(data$response)) / nrow(data) * 100, 1), "%)"),  
    "",
    
    # PFS
    paste(sum(data$event_occurred_pfs == 0, na.rm = TRUE), " (",
          round(sum(data$event_occurred_pfs == 0, na.rm = TRUE) / nrow(data) * 100, 1), "%)"), 
    paste(sum(data$event_occurred_pfs == 1, na.rm = TRUE), "(",
          round(sum(data$event_occurred_pfs == 1, na.rm = TRUE) / nrow(data) * 100, 1), "%)"),
    paste(sum(is.na(data$event_occurred_pfs)), " (",
          round(sum(is.na(data$event_occurred_pfs)) / nrow(data) * 100, 1), "%)"), 
    paste(round(median(data$survival_time_pfs, na.rm = TRUE), 1)),  # Median PFS
    round(IQR(data$survival_time_pfs, na.rm = TRUE), 1),              # IQR PFS

    "",                                       
    
    # 6. OS placeholder
    paste(sum(data$event_occurred_os == 0, na.rm = TRUE), " (",
          round(sum(data$event_occurred_os == 0, na.rm = TRUE) / nrow(data) * 100, 1), "%)"),  # Alive percentage for OS
    paste(sum(data$event_occurred_os == 1, na.rm = TRUE), " (",
          round(sum(data$event_occurred_os == 1, na.rm = TRUE) / nrow(data) * 100, 1), "%)"),  # Dead percentage for OS
    paste(sum(is.na(data$event_occurred_os)), " (",
          round(sum(is.na(data$event_occurred_os)) / nrow(data) * 100, 1), "%)"),  # NA percentage for OS

    paste(round(median(data$survival_time_os, na.rm = TRUE), 1)),  # Median OS
    round(IQR(data$survival_time_os, na.rm = TRUE), 1)              # IQR OS
  )
  
  return(All_DATA)
}



All_DATA = generate_all_data(combined_clin)
Data_with_Sex = generate_all_data(combined_clin_sex)

# Create the final dataframe with both All Data and Data with Sex
final_df <- data.frame(
  Characteristic = Characteristic,
  All_DATA = generate_all_data(combined_clin),
  Data_with_Sex = generate_all_data(combined_clin_sex)
)

# TODO: Rmeove less than 20 and also IO+ chemo and IO+ targeted delete! "Lymph_node"

# Display the final dataframe with a caption
kable(final_df, caption = "Summary of All Data and Data with Sex", col.names = c("Characteristic", "All Data", "Data with Sex"))


#write.csv(final_df, "~/BHK LAB/Ecotyper/outputs/Summary of All Data and Data with Sex.csv", row.names = TRUE)


```
## Detailed clinical characteristics of the IO datasets

processes clinical data from multiple datasets, extracts key characteristics, and formats them into a summary table.

```{r Table 2}

results <- list()

for (data_path in data_files) {
    # Load the data object
    obj <- load(data_path)  
    se <- get(obj)  
    
    # Extract clinical data
    clin <- data.frame(colData(se))
    
    # Base study name
    base_name <- tools::file_path_sans_ext(basename(data_path))
    study_parts <- unlist(strsplit(base_name, "__"))
    study_name <- study_parts[1]
    
    # Number of samples
    n_samples <- nrow(clin)
    
    # Expression type: Adjust the condition based on actual data
    expression_type <- ifelse(any(clin$rna == "rnaseq"), "TPM", ifelse(any(clin$rna == "microarray"), "log2gNor", NA))
    
    # Age calculations
    clin$age <- as.numeric(as.character(clin$age))
    valid_ages <- na.omit(clin$age)
    age_median <- if(length(valid_ages) > 0) round(as.numeric(median(valid_ages, na.rm = TRUE)), 1) else NA
    age_iqr <- if(length(valid_ages) > 1) round(IQR(valid_ages, na.rm = TRUE), 1) else NA
    age_na <- sum(is.na(clin$age))
  

    
    # Sex calculations
    n_female <- sum(clin$sex == "F", na.rm = TRUE)
    n_male <- sum(clin$sex == "M", na.rm = TRUE)
    sex_na <- sum(is.na(clin$sex))
    
    # Cancer type
    cancer_type <- study_parts[2]  # Assuming the second part is the cancer type
    
    # Treatment type
    treatment_type <- paste(unique(clin$treatment), collapse = ", ")
    
    # Response counts
    no_response <- sum(clin$response == "NR", na.rm = TRUE)
    response <- sum(clin$response == "R", na.rm = TRUE)
    
    # PFS counts
    no_progression <- sum(clin$event_occurred_pfs == 0, na.rm = TRUE)
    progression <- sum(clin$event_occurred_pfs == 1, na.rm = TRUE)
    pfs_median <- if (!all(is.na(clin$event_occurred_pfs))) round(median(na.omit(clin$event_occurred_pfs), na.rm = TRUE), 1) else NA
    pfs_iqr <- if (!all(is.na(clin$event_occurred_pfs))) round(IQR(na.omit(clin$event_occurred_pfs), na.rm = TRUE), 1) else NA
    
    # OS counts
    alive <- sum(clin$event_occurred_os == 0, na.rm = TRUE)
    dead <- sum(clin$event_occurred_os == 1, na.rm = TRUE)
    os_median <- if (!all(is.na(clin$event_occurred_os))) round(median(na.omit(clin$event_occurred_os), na.rm = TRUE), 1) else NA
    os_iqr <- if (!all(is.na(clin$event_occurred_os))) round(IQR(na.omit(clin$event_occurred_os), na.rm = TRUE), 1) else NA
    
    # Compile the results for this study
    results[[base_name]] <- data.frame(Study = study_name, N = n_samples, Expression = expression_type, 
                                       Age_Median = age_median, Age_IQR = age_iqr, Age_NA = age_na, 
                                       Female = n_female, Male = n_male, Sex_NA = sex_na, 
                                       Cancer_Type = cancer_type, Treatment_Type = treatment_type, 
                                       No_Response = no_response, Response = response, 
                                       No_Progression = no_progression, Progressed = progression, 
                                       PFS_Median = pfs_median, PFS_IQR = pfs_iqr, 
                                       Alive = alive, Dead = dead, OS_Median = os_median, OS_IQR = os_iqr,
                                       stringsAsFactors = FALSE)
}

# Convert results to a data frame
results_df <- do.call(rbind, results)

# Display the results in a nicely formatted table
kable(results_df, caption = "Detailed clinical characteristics of the IO datasets")

#write.csv(results_df, "~/BHK LAB/Ecotyper/outputs/Detailed_clinical_characteristics_of_the_IO_datasets.csv", row.names = FALSE)

# TODO: delete some studies ( per-cancer and per-sex , per-treatmnets): if <50 or 20 delete this.
# TODO: only PRE patient, Gide and Riaz and Abacus etc.( use  theprevious dataset)
# keep rnaseq ( elgibility : only rna seq , 20 atleast and rnaseq only)

```

## Table S4: Detailed clinical characteristics of the IO dataset stratify by sex 																									
Processes clinical data stratified by sex from multiple datasets, extracts key characteristics, and formats them into a summary table.

```{r Table 3}


results <- list()

for (data_path in data_sex_files) {
    # Load the data object
    obj <- load(data_path)  
    clin <- get(obj) 
    
    # Base study name
    base_name <- tools::file_path_sans_ext(basename(data_path))
    study_parts <- unlist(strsplit(base_name, "__"))
    study_name <- study_parts[1]
    
    # Number of samples
    n_samples <- nrow(clin)
    
    # Expression type: Adjust the condition based on actual data
    expression_type <- ifelse(any(clin$rna == "rnaseq"), "TPM", ifelse(any(clin$rna == "microarray"), "log2gNor", NA))
    
    # Age calculations
    valid_ages <- na.omit(clin$age)
    age_median <- if(length(valid_ages) > 0) round(as.numeric(median(valid_ages, na.rm = TRUE)), 1) else NA
    age_iqr <- if(length(valid_ages) > 1) round(IQR(valid_ages, na.rm = TRUE), 1) else NA
    age_na <- sum(is.na(clin$age))
    
    # Sex calculations
    n_female <- sum(clin$sex == "F", na.rm = TRUE)
    n_male <- sum(clin$sex == "M", na.rm = TRUE)
    sex_na <- sum(is.na(clin$sex))
    
    # Cancer type
    cancer_type <- study_parts[2]  # Assuming the second part is the cancer type
    
    # Treatment type
    treatment_type <- paste(unique(clin$treatment), collapse = ", ")
    
    # Response counts
    no_response <- sum(clin$response == "NR", na.rm = TRUE)
    response <- sum(clin$response == "R", na.rm = TRUE)
    
    # PFS counts
    no_progression <- sum(clin$event_occurred_pfs == 0, na.rm = TRUE)
    progression <- sum(clin$event_occurred_pfs == 1, na.rm = TRUE)
    pfs_median <- if (!all(is.na(clin$event_occurred_pfs))) round(median(na.omit(clin$event_occurred_pfs), na.rm = TRUE), 1) else NA
    pfs_iqr <- if (!all(is.na(clin$event_occurred_pfs))) round(IQR(na.omit(clin$event_occurred_pfs), na.rm = TRUE), 1) else NA
    
    # OS counts
    alive <- sum(clin$event_occurred_os == 0, na.rm = TRUE)
    dead <- sum(clin$event_occurred_os == 1, na.rm = TRUE)
    os_median <- if (!all(is.na(clin$event_occurred_os))) round(median(na.omit(clin$event_occurred_os), na.rm = TRUE), 1) else NA
    os_iqr <- if (!all(is.na(clin$event_occurred_os))) round(IQR(na.omit(clin$event_occurred_os), na.rm = TRUE), 1) else NA
    
    # Compile the results for this study
    results[[base_name]] <- data.frame(Study = study_name, N = n_samples, Expression = expression_type, 
                                       Age_Median = age_median, Age_IQR = age_iqr, Age_NA = age_na, 
                                       Female = n_female, Male = n_male, Sex_NA = sex_na, 
                                       Cancer_Type = cancer_type, Treatment_Type = treatment_type, 
                                       No_Response = no_response, Response = response, 
                                       No_Progression = no_progression, Progressed = progression, 
                                       PFS_Median = pfs_median, PFS_IQR = pfs_iqr, 
                                       Alive = alive, Dead = dead, OS_Median = os_median, OS_IQR = os_iqr,
                                       stringsAsFactors = FALSE)
}

# Convert results to a data frame
results_df <- do.call(rbind, results)

# Display the results in a nicely formatted table
kable(results_df, caption = "Detailed clinical characteristics of the IO dataset stratify by sex ")

#write.csv(results_df, "~/BHK LAB/Ecotyper/outputs/Detailed clinical characteristics of the IO dataset stratify by sex s.csv", row.names = FALSE)


```

## Deconvolution Methods, Starting with Jung Lung PD-1/PD-L1 Dataset

Begin with one dataset, focusing on responders (R) vs. non-responders (NR).
Use a function to retrieve multi-assay data and analyze immune cell proportions based on expression data.

Start with CIBERSORT to identify tumor microenvironment (TME) signatures, such as stromal or immune cells.
The deconvolution method (CIBERSORT) will show cell type distribution based on the data.
The model used here is a linear model → visualized using a heatmap.

Also, apply other deconvolution methods like xCell, EPIC, TIMER, and QUANTISEQ (not a priority).

Focus on the Jung Lung (PD-1/PD-L1) dataset and analyze the pattern of TME for responders (R).


```{r Deconvulation method for Jun Jun datastes}

# Load the package
library(immunedeconv)
library(pheatmap)
# TODO: package "IOBR"- CRAN.

dat_icb <- get(load("~/BHK lab/Ecotyper/data/ICB_Jung__Lung__PD-(L)1.rda"))
expr_Jung <- data.frame(assays(dat_icb))
expr_Jung$group <- NULL
expr_Jung$group_name <- NULL

clin_jung <- data.frame(colData(dat_icb))

expr_Jung <- 2^expr_Jung - 0.001

# Perform deconvolution using quanTIseq
result_quanTIseq <- deconvolute(expr_Jung, method = "quantiseq")

# View the results
head(result_quanTIseq)

# Convert results to a matrix for visualization
result_matrix <- as.matrix(result_quanTIseq[, -1])  # Exclude the first column (cell type names)

# Create a heatmap
pheatmap(result_matrix, cluster_rows = TRUE, cluster_cols = TRUE, display_numbers = TRUE, 
         main = "quanTIseq Deconvolution Heatmap")


# TODO: Positive expression. For all expression data, ensure it is on the same scale using log2(TPM + 1).
# TODO: Create a table specifying the normalized expression that is accepted:
# 1. Columns (e.g., counts, RNA-seq, TPM, etc.)
# 2. Generate Table:
#      - Method: Find available deconvolution methods in the "IOBR" package and include the following:
#      - Accepted expression and normalization (e.g., counts, RNA-seq, TPM, etc.)
#      - PubMed ID for each method
#      - Type of cell (e.g., CIBERSORT --> Immune cells, other combinations)
#      - Number of cells for each method.


# TODO: Perfrom each method+ Create a heatmap for each deconvolution method for a dataset.
# TODO: Write a separate script for comparing deconvolution methods against data.

```
