---
title: "Data processing"
author: "Nasim Bondar Sahebi"
date: "2023-11-07"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## load libraries

```{r libraries}
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(dplyr)
library(knitr)
library(kableExtra)
library(tools)
```


## Data Loading and Preparation

This section lists and displays all `.rda` files from the `data` and `data_sex` directories, representing 47 study cohorts.


```{r  all data}

# Set base directory
Base_dir <- "~/BHK lab/EcoTyper_Project/"

# List .rda files in both directories
data_files <- list.files(path = file.path(Base_dir, "data_all"), pattern = "\\.rda$", full.names = TRUE)
data_sex_files <- list.files(path = file.path(Base_dir, "data_sex"), pattern = "\\.rda$", full.names = TRUE)

data_files_df <- data.frame(Files = basename(data_files))

kable(data_files_df, caption = "All studies", col.names = "Cohorts") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")
```

## Combine all Datasets

This section loads and combines clinical datasets from the `data` and `data_sex` directories. Key steps include:

- Listing `.rda` files from both directories.
- Extracting and merging clinical data based on common columns.
- Creating unique study and patient identifiers.
- Displaying combined clinical data in tables.

The final datasets are saved as `.rds` files for future use.


```{r Combine clinicaldata 1 }

# List .rda files in both directories
data_files <- list.files(path = file.path(Base_dir, "data_all"), pattern = "\\.rda$", full.names = TRUE)
data_sex_files <- list.files(path = file.path(Base_dir, "data_sex"), pattern = "\\.rda$", full.names = TRUE)

# Initialize lists for storing column names and clin dataframes for data_files
column_list <- list()
clin_list <- list()

for (data_path in data_files) {
    obj <- load(data_path)  
    se <- get(obj)
    
    clin <- as.data.frame(colData(se))  
    
    # Set row names with base name and index
    base_name <- tools::file_path_sans_ext(basename(data_path))
    rownames(clin) <- paste0(base_name, "_", seq_len(nrow(clin)))
    
    column_list[[data_path]] <- colnames(clin)  # Store column names
    clin_list[[data_path]] <- clin  # Store clinical data
}

# Identify common columns across all datasets
common_columns <- Reduce(intersect, column_list)

# Bind all clin dataframes using the common columns
combined_clin <- do.call(rbind, lapply(clin_list, function(clin) {
    clin[, common_columns, drop = FALSE]  # Subset to common columns
}))

# Create study and patient identifiers
combined_clin$study <- gsub("\\..*","", basename(rownames(combined_clin)))
combined_clin$study_patient <- paste0(combined_clin$study, "_", combined_clin$patientid)
rownames(combined_clin) <- combined_clin$study_patient

# Repeat for data_sex_files
data_sex_files_df <- data.frame(Files = basename(data_sex_files)) # List all data_sex .rda files

# Initialize lists for storing column names and clin dataframes for data_sex
column_sex_list <- list()
clin_sex_list <- list()

for (data_sex_path in data_sex_files) {
    obj <- load(data_sex_path)  
    clin_sex <- get(obj)
    
    # Set row names with base name and index
    base_name <- tools::file_path_sans_ext(basename(data_sex_path))
    rownames(clin_sex) <- paste0(base_name, "_", seq_len(nrow(clin_sex)))
    
    column_sex_list[[data_sex_path]] <- colnames(clin_sex)  # Store column names
    clin_sex_list[[data_sex_path]] <- clin_sex  # Store clinical data
}

# Identify common columns across all data_sex datasets
common_sex_columns <- Reduce(intersect, column_sex_list)

# Bind all clin_sex dataframes using the common columns
combined_clin_sex <- do.call(rbind, lapply(clin_sex_list, function(clin_sex) {
    clin_sex[, common_sex_columns, drop = FALSE]  # Subset to common columns
}))

# Create study and patient identifiers for data_sex
combined_clin_sex <- combined_clin_sex[!is.na(combined_clin_sex$patientid), ]
combined_clin_sex$study <- gsub("\\..*","", basename(rownames(combined_clin_sex)))
combined_clin_sex$study_patient <- paste0(combined_clin_sex$study, "_", combined_clin_sex$patientid)
rownames(combined_clin_sex) <- combined_clin_sex$study_patient

# Display the combined_clin table with scrolling enabled
kable(head(combined_clin), caption = "Summary of Combined Clinical Data (Data Directory)", 
      col.names = colnames(combined_clin), format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")

```

## Summary of All Data and Data with Sex

This section generates a summary table of clinical characteristics from the combined clinical datasets. The table includes the number of studies, patients, age (with median and IQR), missing values (NA %), sex distribution, cancer types, treatments, response rates, and survival outcomes (PFS and OS). The function `generate_all_data()` processes the data and computes these metrics for both the full dataset and the dataset with sex-specific information.

### Exclusion Criteria:

- Remove rows with the following treatments: `"IO+chemo"`, `"IO+targeted"`, `"chemo"`, `"chemo+targeted"`, and `"targeted"`.
- Remove cancer types with fewer than 20 patients.
- Exclude the cancer type `"Lymph_node"`.
- Only include patients from the studies `"Gide"`, `"Riaz"`, and `"Abacus"`, and only include `"PRE"` samples.
- Only keep patients with RNA-seq data (exclude microarray patients)
- only keeping PRE samples 


```{r Table 1 }

# Remove rows with treatments "IO+chemo", "IO+targeted", "chemo", "chemo+targeted", and "targeted"
combined_clin <- combined_clin[!combined_clin$treatment %in% c("IO+chemo", "IO+targeted", "chemo", "chemo+targeted", "targeted"), ]
combined_clin_sex <- combined_clin_sex[!combined_clin_sex$treatment %in% c("IO+chemo", "IO+targeted", "chemo", "chemo+targeted", "targeted"), ]


# Remove cancer types with less than 20 patients
cancer_types_less_20 <- names(table(combined_clin$cancer_type)[table(combined_clin$cancer_type) < 20])
combined_clin <- combined_clin[!combined_clin$cancer_type %in% cancer_types_less_20, ]
combined_clin_sex <- combined_clin_sex[!combined_clin_sex$cancer_type %in% cancer_types_less_20, ]

# Remove "Lymph_node" cancer type
combined_clin <- combined_clin[combined_clin$cancer_type != "Lymph_node", ]
combined_clin_sex <- combined_clin_sex[combined_clin_sex$cancer_type != "Lymph_node", ]

# Subset only patients with RNA-seq data (exclude microarray patients)
combined_clin <- combined_clin[combined_clin$rna == "rnaseq", ] #186 patients removed
combined_clin_sex <- combined_clin_sex[combined_clin_sex$rna == "rnaseq", ]


#saveRDS(combined_clin, file = file.path("~/BHK LAB/EcoTyper_Project//outputs/", "combined_clin.rds"))

#saveRDS(combined_clin_sex, file = file.path("~/BHK LAB/EcoTyper_Project//outputs/", "combined_clin_sex.rds"))

treatments <- unique(combined_clin_sex$treatment)
cancer_types <- unique(combined_clin_sex$cancer_type)
                          
Characteristic <- c("N° studies", "N° patients", "Age year", "Median (IQR)", "NA (%)", "Sex", "Female (%)", "Male (%)", "NA (%)", "Cancer Type",paste0(cancer_types, " (%)"), "Treatment", paste0(treatments, " (%)"), "Response", "No-response (%)", "Response (%)", "NA (%)", "PFS", "No-progression (%)", "Progressed (%)", "NA (%)", "Median PFS", "IQR PFS", "OS", "Alive (%)", "Dead (%)", "NA (%)", "Median OS", "IQR OS")


# Function to extract the base of the study name, ignoring the "_F" and "_M" ( for counting data sex with same study name)
get_study_base <- function(study_name) {
  return(sub("_[FM]$", "", study_name))
}

# Generate data function with only the common treatments and cancer types
generate_all_data <- function(data) {
  
  data$study_base <- sapply(data$study, get_study_base)
   
  All_DATA <- c(
    length(unique(data$study_base)),    # N° studies          
    length(data$patientid),                    # N° patients          
    "",                                       # Age year placeholder
    
    # 1. Age year placeholder
    paste(median(as.numeric(data$age), na.rm = TRUE), " (", min(as.numeric(data$age), na.rm = TRUE), "-", max(as.numeric(data$age), na.rm = TRUE), ")", sep = ""),
    paste(sum(is.na(data$age)), " (", round(sum(is.na(data$age)) / nrow(data) * 100, 1), "%)", sep = ""),
        
    "",                                     
    
    # 2. Sex placeholder
    paste(sum(data$sex == "F", na.rm = TRUE) , " (",round(sum(data$sex == "F", na.rm = TRUE) / nrow(data) * 100, 1), "%)"),  # Female percentage
    paste( sum(data$sex == "M", na.rm = TRUE), " (",round(sum(data$sex == "M", na.rm = TRUE) / nrow(data) * 100, 1), "%)"),  # Male percentage
    paste(sum(is.na(data$sex)) , " (",round(sum(is.na(data$sex)) / nrow(data) * 100, 1), "%)"),  # NA percentage for sex
    "",                                       # Cancer Type placeholder
    
    # 3. Calculate common cancer type percentages
    sapply(cancer_types, function(ct) {
      paste(sum(data$cancer_type == ct), " (", round(sum(data$cancer_type == ct) / nrow(data) * 100, 1), "%)", sep = "")
    }), 
    "",                                       
    
    # 4. Treatment placeholder for common treatments
    sapply(treatments, function(tr) {
      paste(sum(data$treatment == tr), " (", round(sum(data$treatment == tr) / nrow(data) * 100, 1), "%)", sep = "")
    }), 
    "",                                      
    
    # 5. Response placeholder
    paste(sum(data$response == "NR", na.rm = TRUE), " (", round(sum(data$response == "NR", na.rm = TRUE) / nrow(data) * 100, 1), "%)"), 
    paste(sum(data$response == "R", na.rm = TRUE), " (",round(sum(data$response == "R", na.rm = TRUE) / nrow(data) * 100, 1), "%)"),
    paste(sum(is.na(data$response)), "(", round(sum(is.na(data$response)) / nrow(data) * 100, 1), "%)"),  
    "",
    
    # PFS
    paste(sum(data$event_occurred_pfs == 0, na.rm = TRUE), " (",
          round(sum(data$event_occurred_pfs == 0, na.rm = TRUE) / nrow(data) * 100, 1), "%)"), 
    paste(sum(data$event_occurred_pfs == 1, na.rm = TRUE), "(",
          round(sum(data$event_occurred_pfs == 1, na.rm = TRUE) / nrow(data) * 100, 1), "%)"),
    paste(sum(is.na(data$event_occurred_pfs)), " (",
          round(sum(is.na(data$event_occurred_pfs)) / nrow(data) * 100, 1), "%)"), 
    paste(round(median(data$survival_time_pfs, na.rm = TRUE), 1)),  # Median PFS
    round(IQR(data$survival_time_pfs, na.rm = TRUE), 1),              # IQR PFS

    "",                                       
    
    # 6. OS placeholder
    paste(sum(data$event_occurred_os == 0, na.rm = TRUE), " (",
          round(sum(data$event_occurred_os == 0, na.rm = TRUE) / nrow(data) * 100, 1), "%)"),  # Alive percentage for OS
    paste(sum(data$event_occurred_os == 1, na.rm = TRUE), " (",
          round(sum(data$event_occurred_os == 1, na.rm = TRUE) / nrow(data) * 100, 1), "%)"),  # Dead percentage for OS
    paste(sum(is.na(data$event_occurred_os)), " (",
          round(sum(is.na(data$event_occurred_os)) / nrow(data) * 100, 1), "%)"),  # NA percentage for OS

    paste(round(median(data$survival_time_os, na.rm = TRUE), 1)),  # Median OS
    round(IQR(data$survival_time_os, na.rm = TRUE), 1)              # IQR OS
  )
  
  return(All_DATA)
}


All_DATA = generate_all_data(combined_clin)
Data_with_Sex = generate_all_data(combined_clin_sex)

# Create the final dataframe with both All Data and Data with Sex
final_df <- data.frame(
  Characteristic = Characteristic,
  All_DATA = generate_all_data(combined_clin),
  Data_with_Sex = generate_all_data(combined_clin_sex)
)


kable(final_df, caption = "Summary of All Data and Data with Sex", 
      col.names = c("Characteristic", "All Data", "Data with Sex")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")


# write.csv(final_df, "~/BHK LAB/EcoTyper_Project/outputs/Summary of All Data and Data with Sex.csv", row.names = TRUE)


```
# Studies that we are focusing on after the exclusion

```{r Studies after exclusion}

kable(unique(combined_clin$study), caption = "All studies", col.names = "Cohorts") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")

```


## Detailed clinical characteristics of the IO datasets

processes clinical data from multiple datasets, extracts key characteristics, and formats them into a summary table.

```{r Table 2}

# Initialize the results list
results <- list()

# Loop through unique studies in combined_clin
for (study_name in unique(combined_clin$study)) {
  
  # Subset the data for the current study
  study_data <- combined_clin[combined_clin$study == study_name, ]
  
  # Number of samples
  n_samples <- nrow(study_data)
  
  # Expression type: Adjust based on actual data
  expression_type <- ifelse(any(study_data$rna == "rnaseq"), "TPM", ifelse(any(study_data$rna == "microarray"), "log2gNor", NA))
  
  # Age calculations
  study_data$age <- as.numeric(as.character(study_data$age))
  valid_ages <- na.omit(study_data$age)
  age_median <- if(length(valid_ages) > 0) round(median(valid_ages, na.rm = TRUE), 1) else NA
  age_iqr <- if(length(valid_ages) > 1) round(IQR(valid_ages, na.rm = TRUE), 1) else NA
  age_na <- sum(is.na(study_data$age))
  
  # Sex calculations
  n_female <- sum(study_data$sex == "F", na.rm = TRUE)
  n_male <- sum(study_data$sex == "M", na.rm = TRUE)
  sex_na <- sum(is.na(study_data$sex))
  
  # Cancer type (assuming the cancer type is consistent within the study)
  cancer_type <- unique(study_data$cancer_type)
  
  # Treatment type
  treatment_type <- paste(unique(study_data$treatment), collapse = ", ")
  
  # Response counts
  no_response <- sum(study_data$response == "NR", na.rm = TRUE)
  response <- sum(study_data$response == "R", na.rm = TRUE)
  
  # PFS counts
  no_progression <- sum(study_data$event_occurred_pfs == 0, na.rm = TRUE)
  progression <- sum(study_data$event_occurred_pfs == 1, na.rm = TRUE)
  pfs_median <- if (!all(is.na(study_data$survival_time_pfs))) round(median(na.omit(study_data$survival_time_pfs), na.rm = TRUE), 1) else NA
  pfs_iqr <- if (!all(is.na(study_data$survival_time_pfs))) round(IQR(na.omit(study_data$survival_time_pfs), na.rm = TRUE), 1) else NA
  
  # OS counts
  alive <- sum(study_data$event_occurred_os == 0, na.rm = TRUE)
  dead <- sum(study_data$event_occurred_os == 1, na.rm = TRUE)
  os_median <- if (!all(is.na(study_data$survival_time_os))) round(median(na.omit(study_data$survival_time_os), na.rm = TRUE), 1) else NA
  os_iqr <- if (!all(is.na(study_data$survival_time_os))) round(IQR(na.omit(study_data$survival_time_os), na.rm = TRUE), 1) else NA
  
  # Compile the results for this study
  results[[study_name]] <- data.frame(
    Study = study_name, 
    N = n_samples, 
    Expression = expression_type, 
    Age_Median = age_median, 
    Age_IQR = age_iqr, 
    Age_NA = age_na, 
    Female = n_female, 
    Male = n_male, 
    Sex_NA = sex_na, 
    Cancer_Type = cancer_type, 
    Treatment_Type = treatment_type, 
    No_Response = no_response, 
    Response = response, 
    No_Progression = no_progression, 
    Progressed = progression, 
    PFS_Median = pfs_median, 
    PFS_IQR = pfs_iqr, 
    Alive = alive, 
    Dead = dead, 
    OS_Median = os_median, 
    OS_IQR = os_iqr,
    stringsAsFactors = FALSE
  )
}

# Combine all results into a single dataframe
results_df <- do.call(rbind, results)

kable(results_df, caption = "Detailed clinical characteristics of the IO datasets") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")

#write.csv(results_df, "~/BHK LAB/EcoTyper_Project/outputs/Detailed_clinical_characteristics_of_the_IO_datasets.csv", row.names = FALSE)


```

## Table S4: Detailed clinical characteristics of the IO dataset stratify by sex 																									
Processes clinical data stratified by sex from multiple datasets, extracts key characteristics, and formats them into a summary table.

```{r Table 3}

# Initialize the results list for combined_clin_sex
results_sex <- list()

# Loop through unique studies in combined_clin_sex
for (study_name in unique(combined_clin_sex$study)) {
  
  # Subset the data for the current study
  study_data <- combined_clin_sex[combined_clin_sex$study == study_name, ]
  
  # Number of samples
  n_samples <- nrow(study_data)
  
  # Expression type: Adjust based on actual data
  expression_type <- ifelse(any(study_data$rna == "rnaseq"), "TPM", ifelse(any(study_data$rna == "microarray"), "log2gNor", NA))
  
  # Age calculations
  study_data$age <- as.numeric(as.character(study_data$age))
  valid_ages <- na.omit(study_data$age)
  age_median <- if(length(valid_ages) > 0) round(median(valid_ages, na.rm = TRUE), 1) else NA
  age_iqr <- if(length(valid_ages) > 1) round(IQR(valid_ages, na.rm = TRUE), 1) else NA
  age_na <- sum(is.na(study_data$age))
  
  # Sex calculations
  n_female <- sum(study_data$sex == "F", na.rm = TRUE)
  n_male <- sum(study_data$sex == "M", na.rm = TRUE)
  sex_na <- sum(is.na(study_data$sex))
  
  # Cancer type (assuming the cancer type is consistent within the study)
  cancer_type <- unique(study_data$cancer_type)
  
  # Treatment type
  treatment_type <- paste(unique(study_data$treatment), collapse = ", ")
  
  # Response counts
  no_response <- sum(study_data$response == "NR", na.rm = TRUE)
  response <- sum(study_data$response == "R", na.rm = TRUE)
  
  # PFS counts
  no_progression <- sum(study_data$event_occurred_pfs == 0, na.rm = TRUE)
  progression <- sum(study_data$event_occurred_pfs == 1, na.rm = TRUE)
  pfs_median <- if (!all(is.na(study_data$survival_time_pfs))) round(median(na.omit(study_data$survival_time_pfs), na.rm = TRUE), 1) else NA
  pfs_iqr <- if (!all(is.na(study_data$survival_time_pfs))) round(IQR(na.omit(study_data$survival_time_pfs), na.rm = TRUE), 1) else NA
  
  # OS counts
  alive <- sum(study_data$event_occurred_os == 0, na.rm = TRUE)
  dead <- sum(study_data$event_occurred_os == 1, na.rm = TRUE)
  os_median <- if (!all(is.na(study_data$survival_time_os))) round(median(na.omit(study_data$survival_time_os), na.rm = TRUE), 1) else NA
  os_iqr <- if (!all(is.na(study_data$survival_time_os))) round(IQR(na.omit(study_data$survival_time_os), na.rm = TRUE), 1) else NA
  
  # Compile the results for this study
  results_sex[[study_name]] <- data.frame(
    Study = study_name, 
    N = n_samples, 
    Expression = expression_type, 
    Age_Median = age_median, 
    Age_IQR = age_iqr, 
    Age_NA = age_na, 
    Female = n_female, 
    Male = n_male, 
    Sex_NA = sex_na, 
    Cancer_Type = cancer_type, 
    Treatment_Type = treatment_type, 
    No_Response = no_response, 
    Response = response, 
    No_Progression = no_progression, 
    Progressed = progression, 
    PFS_Median = pfs_median, 
    PFS_IQR = pfs_iqr, 
    Alive = alive, 
    Dead = dead, 
    OS_Median = os_median, 
    OS_IQR = os_iqr,
    stringsAsFactors = FALSE
  )}
  
# Combine all results into a single dataframe
results_df_sex <- do.call(rbind, results_sex)

# Display 
kable(results_df_sex, caption = "Detailed clinical characteristics of the IO datasets") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")


# Write to CSV
write.csv(results_df_sex, "~/BHK LAB/EcoTyper_Project/outputs/Detailed_clinical_characteristics_of_the_IO_datasets_combined_clin_sex.csv", row.names = FALSE)

```


