---
title: "Deconvolution Porcess"
author: "Nasim Bondar Sahebi"
date: "2023-10-10"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## load libraries

```{r libraries}
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(immunedeconv)
library(ComplexHeatmap)
library(circlize)
library(knitr)
library(kableExtra)
library(dplyr)
library(IOBR)
#library(tools)
```

## Deconvolution Methods

### Overview of Dataset and Clinical Characteristics

#### Inclusion Criteria for Dataset Analysis:

- **Treatments**: Patients treated with `"PD-1/PD-L1"`, `"IO+combo"`, or `"CTLA4"` were included.
- **Cancer Types**: Only cancer types with at least 20 patients were considered. Non-solid tumors were prioritized, while solid tumors like `"Lymph_node"` were excluded.
- **Sample Types**: Focused on `"PRE"` treatment samples to assess baseline characteristics before therapy.
- **RNA-seq Data**: Only patients with RNA-seq data were included; those with microarray data were excluded.
- **Data Normalization**: RNA-seq data was normalized using `log2(TPM + 0.001)` to ensure comparable expression levels across patients.

#### Summary of Clinical Characteristics:

- **Number of Studies**: 38 (All Data), 51 (Data with Sex).
- **Number of Patients**: 2,178 (All Data), 1,758 (Data with Sex).

##### Sex Distribution:

- **Female**: 32.6% (All Data), 40.3% (Data with Sex).
- **Male**: 48.2% (All Data), 59.7% (Data with Sex).
- **Sex Not Available**: 19.3% (All Data), 0% (Data with Sex).

##### Cancer Types:

- **Lung Cancer**: 33.9% (All Data), 36.6% (Data with Sex).
- **Melanoma**: 25.4% (All Data), 23.2% (Data with Sex).
- **Bladder Cancer**: 16.9% (All Data), 16.2% (Data with Sex).
- **Other Types**: Includes smaller proportions of kidney, brain, ureteral, pancreas, and esophageal cancers.

##### Treatment Types:

- **PD-1/PD-L1**: 90.4%.
- **IO+combo**: 5.6%.
- **CTLA4**: 4%.

##### Treatment Response:

- **Non-Responders**: 55.2% (All Data), 55.6% (Data with Sex).
- **Responders**: 25.2% (All Data), 25.7% (Data with Sex).

##### Progression-Free Survival (PFS):

- **Median PFS**: 3.1 months.
- **IQR PFS**: 7.9 months (All Data), 8.6 months (Data with Sex).

##### Overall Survival (OS):

- **Median OS**: 12.2 months (All Data), 12.1 months (Data with Sex).
- **IQR OS**: 16.8 months (All Data), 16.7 months (Data with Sex).

```{r Deconvulation method for Jun Jun datastes}
# Set base directory
Base_dir <- "~/BHK lab/EcoTyper_Project/"

# Load and display Summary and Clinical datasets
kable(read.csv(file.path(Base_dir, "outputs", "Summary of All Data and Data with Sex.csv")), 
      caption = "Summary of All Data and Data with Sex") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")

kable(read.csv(file.path(Base_dir, "outputs", "Detailed_clinical_characteristics_of_the_IO_datasets.csv")), 
      caption = "Detailed Clinical Characteristics of the IO Datasets") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")

```


### 2. Generate a Table of IOBR Deconvolution Methods

Table summarizing the deconvolution methods in the  [IOBR package](https://github.com/IOBR/IOBR). Include the following columns:

1. **Method**: List the available deconvolution methods.
2. **Accepted Expression and Normalization**: Specify accepted expression types (e.g., counts, RNA-seq, TPM, etc.).
3. **PubMed ID**: Include the PubMed ID or DOI link for each method.
4. **Type of Cell**: Indicate the type of cells targeted by each method (e.g., CIBERSORT --> Immune cells).
5. **Number of Cells**: Provide the number of cell types for each method.


```{r IBOR Deconvulation methods}

methods_table <- data.frame(
  Method = c(
    "[CIBERSORT](https://cibersort.stanford.edu/)",
    "[EPIC](https://gfellerlab.shinyapps.io/EPIC_1-1/)",
    "[MCPcounter](https://github.com/ebecht/MCPcounter/blob/master/Source/License)",
    "[xCell](https://github.com/dviraran/xCell/blob/master/DESCRIPTION)",
    "[ESTIMATE](https://bioinformatics.mdanderson.org/public-software/estimate/)",
    "[TIMER](http://cistrome.org/TIMER/download.html)",
    "[quanTIseq](http://icbi.at/software/quantiseq/doc/index.html)",
    "[IPS](https://github.com/icbi-lab/Immunophenogram/blob/master/LICENSE)"
  ),
  Accepted_Expression_Normalization = c(
    "Counts, RNA-seq, TPM, FPKM",
    "RNA-seq, TPM, FPKM",
    "RNA-seq, TPM, FPKM",
    "RNA-seq, TPM, FPKM, Counts",
    "RNA-seq, FPKM",
    "RNA-seq, TPM",
    "RNA-seq, TPM",
    "RNA-seq, TPM"
  ),
  PubMed_ID = c(
    "[PMID: 25977888](https://doi.org/10.1038/nmeth.3337)",
    "[PMID: 28783075](https://doi.org/10.7554/eLife.26476)",
    "[PMID: 27798624](https://doi.org/10.1186/s13059-016-1070-5)",
    "[PMID: 29141660](https://doi.org/10.1186/s13059-017-1349-1)",
    "[PMID: 24113773](https://doi.org/10.1038/ncomms3612)",
    "[PMID: 27423106](https://doi.org/10.1186/s13059-016-1028-7)",
    "[PMID: 30992056](https://doi.org/10.1186/s13073-019-0638-6)",
    "[PMID: 28052254](https://doi.org/10.1016/j.celrep.2016.12.019)"
  ),
  Cell_Type = c(
    "Immune cells",
    "Cancer cells, Immune cells",
    "Immune cells, Stromal cells",
    "Immune cells, Stromal cells",
    "Tumor purity, Stromal and Immune cells",
    "Immune cells",
    "Tumor immune cells",
    "Immune phenotypes"
  ),
  Number_of_Cells = c(
    "22 immune cell types",
    "Multiple (B cells, T cells, Macrophages)",
    "10 cell types",
    "64 cell types",
    "Not specified",
    "6 immune cell types",
    "10 immune cell types",
    "28 immune cell types"
  ),
  License = c(
    "free for non-commercial use only",
    "free for non-commercial use only (Academic License)",
    "free [(GPL 3.0)](https://github.com/ebecht/MCPcounter/blob/master/Source/License)",
    "free [(GPL 3.0)](https://github.com/dviraran/xCell/blob/master/DESCRIPTION)",
    "free [(GPL 2.0)](https://bioinformatics.mdanderson.org/estimate/)",
    "free [(GPL 2.0)](http://cistrome.org/TIMER/download.html)",
    "free [(BSD)](https://github.com/icbi-lab/immunedeconv/blob/master/LICENSE.md)",
    "free [(BSD)](https://github.com/icbi-lab/Immunophenogram/blob/master/LICENSE)"
  ),
  Citation = c(
    "Newman, A. M., Liu, C. L., Green, M. R., Gentles, A. J., Feng, W., Xu, Y., … Alizadeh, A. A. (2015). Robust enumeration of cell subsets from tissue expression profiles. Nature Methods, 12(5), 453–457. [https://doi.org/10.1038/nmeth.3337](https://doi.org/10.1038/nmeth.3337)",
    "Racle, J., de Jonge, K., Baumgaertner, P., Speiser, D. E., & Gfeller, D. (2017). Simultaneous enumeration of cancer and immune cell types from bulk tumor gene expression data. ELife, 6, e26476. [https://doi.org/10.7554/eLife.26476](https://doi.org/10.7554/eLife.26476)",
    "Becht, E., Giraldo, N. A., Lacroix, L., Buttard, B., Elarouci, N., Petitprez, F., … de Reyniès, A. (2016). Estimating the population abundance of tissue-infiltrating immune and stromal cell populations using gene expression. Genome Biology, 17(1), 218. [https://doi.org/10.1186/s13059-016-1070-5](https://doi.org/10.1186/s13059-016-1070-5)",
    "Aran, D., Hu, Z., & Butte, A. J. (2017). xCell: digitally portraying the tissue cellular heterogeneity landscape. Genome Biology, 18(1), 220. [https://doi.org/10.1186/s13059-017-1349-1](https://doi.org/10.1186/s13059-017-1349-1)",
    "Vegesna R, Kim H, Torres-Garcia W, …, Verhaak R. (2013). Inferring tumour purity and stromal and immune cell admixture from expression data. Nature Communications 4, 2612. [http://doi.org/10.1038/ncomms3612](http://doi.org/10.1038/ncomms3612)",
    "Li, B., Severson, E., Pignon, J.-C., Zhao, H., Li, T., Novak, J., … Liu, X. S. (2016). Comprehensive analyses of tumor immunity: implications for cancer immunotherapy. Genome Biology, 17(1), 174. [https://doi.org/10.1186/s13059-016-1028-7](https://doi.org/10.1186/s13059-016-1028-7)",
    "Finotello, F., Mayer, C., Plattner, C., Laschober, G., Rieder, D., Hackl, H., …, Sopper, S. (2019). Molecular and pharmacological modulators of the tumor immune contexture revealed by deconvolution of RNA-seq data. Genome medicine, 11(1), 34. [https://doi.org/10.1186/s13073-019-0638-6](https://doi.org/10.1186/s13073-019-0638-6)",
    "Charoentong, P., et al. (2017). Pan-cancer Immunogenomic Analyses Reveal Genotype-Immunophenotype Relationships and Predictors of Response to Checkpoint Blockade. Cell Reports 18, 248-262. [https://doi.org/10.1016/j.celrep.2016.12.019](https://doi.org/10.1016/j.celrep.2016.12.019)"
  )
)

# Generate the table using kable
kable(methods_table, format = "markdown", align = "l", caption = "Deconvolution Methods in IOBR Package with License and Citation Links")

# write.csv(methods_table, file = "~/BHK lab/EcoTyper_Project/outputs/deconvolution_method.table.csv", row.names = TRUE)

```

### Summary of Deconvolution Methods and Algorithms
This table provides an overview of the different deconvolution methods, including their processes for computing scores, the algorithms they use, and whether machine learning (ML) is involved in estimating cell proportions across samples.

```{r}

# DF summarizing the deconvolution methods
methods_summary <- data.frame(
  Method = c(
    "CIBERSORT", 
    "EPIC", 
    "MCPcounter", 
    "xCell", 
    "ESTIMATE", 
    "TIMER", 
    "quanTIseq", 
    "IPS"
  ),
  Computation_Process = c(
    "Uses support vector regression (SVR) to estimate the relative fraction of cell types in bulk tissue samples based on gene expression signatures.",
    "Estimates immune and cancer cell fractions using gene expression data combined with reference profiles.",
    "Counts different cell types using marker genes expressed in immune and stromal cells.",
    "Uses signatures derived from RNA-seq data to calculate cell type scores based on relative enrichment.",
    "Calculates tumor purity, immune, and stromal cell infiltration scores from gene expression data.",
    "Uses gene expression signatures to estimate the abundance of tumor-infiltrating immune cells.",
    "Uses reference profiles and RNA-seq data to estimate immune cell fractions in tumor samples.",
    "Immunophenogram-based method that scores immune cell activity using gene expression data."
  ),
  Algorithm = c(
    "Support Vector Regression (SVR)", 
    "Non-negative least squares (NNLS)", 
    "Marker-based counting", 
    "Single sample gene set enrichment analysis (ssGSEA)", 
    "Single sample gene set enrichment analysis (ssGSEA)", 
    "Gene expression signatures", 
    "Linear regression", 
    "Non-linear scoring using immune-related genes"
  ),
  ML_Algorithm = c(
    "Yes",  # SVR is an ML algorithm
    "No",   # EPIC does not use ML in its core process
    "No",   # MCPcounter does not use ML
    "No",   # xCell does not use ML, relies on enrichment scores
    "No",   # ESTIMATE does not use ML, based on enrichment scores
    "No",   # TIMER uses gene signatures, no ML involved
    "No",   # quanTIseq uses linear regression, no ML involved
    "Yes"   # IPS uses non-linear scoring which may involve ML
  )
)


kable(methods_summary, format = "html", caption = "Summary of Deconvolution Methods and Algorithms") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# save as CSV
#write.csv(methods_summary, file = "~/BHK lab/EcoTyper_Project/outputs/deconvolution_methods_summary.csv", row.names = FALSE)

```
### CIBERSORT on **Jung Lung PD-1/PD-L1**

We begin by analyzing the **Jung Lung PD-1/PD-L1** dataset, which contains clinical characteristics of patients who received PD-1/PD-L1 treatment for lung cancer. The dataset has been normalized using RNA-seq data and includes various clinical factors such as age, sex, response to treatment, progression-free survival (PFS), and overall survival (OS).

**Important Note**: Currently, the RNA-seq data is in log2(TPM + 0.001) format. To ensure consistency, we need to convert the expression data to a positive expression range using **log2(TPM + 1)** for all samples, so all expression data is on the same scale.

We will begin the analysis with a focus on **responders (R) vs. non-responders (NR)**. The first step is to retrieve multi-assay data and analyze immune cell proportions based on the expression data.

We'll start with **CIBERSORT** to identify tumor microenvironment (TME) signatures, such as stromal or immune cell types. CIBERSORT uses a linear model to estimate the cell type distribution within the TME, which will be visualized using a heatmap.

For this analysis, we will focus on the **Jung Lung (PD-1/PD-L1)** dataset, specifically examining the TME patterns for responders (R).

**Next Steps**: We will also apply other deconvolution methods, such as xCell, EPIC, TIMER, and QUANTISEQ, though this is not a priority at this stage.


```{r IBOR Deconvolution method for Jung Jung datasets, fig.height=12, fig.width=10}

# Load the specific .rda file (ICB_Jung_Lung_PD-L1)
ICB_jung <- get(load(file.path(Base_dir, "data_all", "ICB_Jung__Lung__PD-(L)1.rda")))

# Extract expression data and convert to a matrix
expr_Jung <- data.frame(assays(ICB_jung))
expr_Jung$group <- NULL
expr_Jung$group_name <- NULL
expr_Jung <- as.matrix(expr_Jung)
annot_Jung <- data.frame(rowData(ICB_jung))

# Extract clinical data
clin_jung <- data.frame(colData(ICB_jung))

# Step 1: Convert expression data from log2(TPM + 0.001) back to TPM
expr_Jung_TPM <- 2^(expr_Jung) - 0.001
expr_Jung <- log2(expr_Jung_TPM + 1)

# Run deconvolution analysis using CIBERSORT
cibersort_results <- deconvo_tme(eset = expr_Jung, method = "cibersort")
cibersort_results <- as.data.frame(cibersort_results)
rownames(cibersort_results) <- colnames(expr_Jung)
cibersort_results$ID <- NULL
colnames(cibersort_results) <- gsub("_CIBERSORT", "", colnames(cibersort_results))


kable(cibersort_results, format = "html", caption = "CIBERSORT Deconvolution Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))


# Correct column names for removal
filtered_cibersort_results <- cibersort_results %>%
select(-`P-value`, -`Correlation`, -`RMSE`)


##1. non scaled version
reorder_index <- order(clin_jung$response)
 filtered_cibersort_results <-  filtered_cibersort_results[reorder_index, ]
clin_jung <- clin_jung[reorder_index, ]

response_annotation <- HeatmapAnnotation(Response = clin_jung$response)

# Create a heatmap with scaled results
h1 <- Heatmap(t(as.matrix( filtered_cibersort_results)), name = "CIBERSORT Results (Not Scaled)", top_annotation  = response_annotation,cluster_rows = F,column_split = clin_jung$response)



## 2. Apply scaling to each row based on min, median, and max
# Define the color scale for each row
common_color_scale <- function(x) {
  data_min <- min(x)
  data_median <- median(x)
  data_max <- max(x)
  colorRamp2(c(data_min, data_median, data_max), c("#2166AC", "#F7F7F7", "#B2182B"))(x)
}

# Apply the custom color scale row by row
h2 <- Heatmap(t(as.matrix(filtered_cibersort_results)), 
              name = "CIBERSORT Results (Min-Median-Max)", 
              top_annotation = response_annotation, 
              cluster_rows = F, 
              column_split = clin_jung$response,
              col = common_color_scale)



## 3.scaled version with scale
# Scale and cap the filtered CIBERSORT results between -4 and 4
scaled_cibersort_results <- as.data.frame(scale(filtered_cibersort_results))
#scaled_cibersort_results <- pmax(pmin(scaled_cibersort_results, 4), -4)

reorder_index <- order(clin_jung$response)
scaled_cibersort_results <- scaled_cibersort_results[reorder_index, ]
clin_jung <- clin_jung[reorder_index, ]

response_annotation <- HeatmapAnnotation(Response = clin_jung$response)

# Create a heatmap with scaled results
h3 <- Heatmap(t(as.matrix(scaled_cibersort_results)), name = "CIBERSORT Results (Scaled)", top_annotation  = response_annotation,cluster_rows = F,column_split = clin_jung$response)

h1
h2
h3# TODO: keep this.1.

#TODO: shared heatmap(by Franoosh)2.

```

```{r IBOR other Deconvolution  for Jung Jung datasets, fig.height=12, fig.width= 11}

# Define a reusable function for creating both scaled and min-median-max heatmaps
create_heatmaps <- function(data, name) {
  # Reorder data based on response
  reorder_index <- order(clin_jung$response)
  data <- data[reorder_index, ]
  clin_jung <- clin_jung[reorder_index, ]

  # Handle missing values: replace NAs with the median of each column
  data[is.na(data)] <- apply(data, 2, median, na.rm = TRUE)

  # Create annotation for the heatmap
  response_annotation <- HeatmapAnnotation(Response = clin_jung$response)

  # 1. Scaled version
  data_scaled <- as.data.frame(scale(data))
  h_scaled <- Heatmap(t(as.matrix(data_scaled)), name = paste(name, "(Scaled)"), 
                      top_annotation = response_annotation, 
                      cluster_rows = FALSE, column_split = clin_jung$response, 
                      col = colorRamp2(c(-4, 0, 4), c("blue", "white", "red")))

  # Draw the scaled heatmap
  draw(h_scaled)

  # 2. Min-Median-Max version with adjusted color scale
  data_min <- min(as.numeric(unlist(data)), na.rm = TRUE)
  data_median <- median(as.numeric(unlist(data)), na.rm = TRUE)
  data_max <- max(as.numeric(unlist(data)), na.rm = TRUE)
  custom_color_scale <- colorRamp2(c(data_min, data_median, data_max), c("#2166AC", "#F7F7F7", "#B2182B"))

  h_min_median_max <- Heatmap(t(as.matrix(data)), name = paste(name, "(Min-Median-Max)"), 
                              top_annotation = response_annotation, 
                              cluster_rows = FALSE, column_split = clin_jung$response, 
                              col = custom_color_scale)

  # Draw the Min-Median-Max heatmap
  draw(h_min_median_max)
}

# Example usage for EPIC, MCPcounter, xCell, and ESTIMATE results

# Run deconvolution analysis and generate heatmaps for EPIC
epic_results <- as.data.frame(deconvo_tme(eset = expr_Jung, method = "epic"))
rownames(epic_results) <- epic_results$ID
epic_results$ID <- NULL
kable(epic_results, format = "html", caption = "EPIC Deconvolution Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
create_heatmaps(epic_results, "EPIC Results")

# Run deconvolution analysis and generate heatmaps for MCPcounter
mcpcounter_results <- as.data.frame(deconvo_tme(eset = expr_Jung, method = "mcpcounter"))
rownames(mcpcounter_results) <- colnames(expr_Jung)
mcpcounter_results$ID <- NULL
kable(mcpcounter_results, format = "html", caption = "MCPcounter Deconvolution Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
create_heatmaps(mcpcounter_results, "MCPcounter Results")

# Run deconvolution analysis and generate heatmaps for xCell
xcell_results <- as.data.frame(deconvo_tme(eset = expr_Jung, method = "xcell"))
rownames(xcell_results) <- colnames(expr_Jung)
xcell_results$ID <- NULL
kable(xcell_results, format = "html", caption = "xCell Deconvolution Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
create_heatmaps(xcell_results, "xCell Results")

# Run deconvolution analysis and generate heatmaps for ESTIMATE
estimate_results <- as.data.frame(deconvo_tme(eset = expr_Jung, method = "estimate"))
rownames(estimate_results) <- colnames(expr_Jung)
estimate_results$ID <- NULL
kable(estimate_results, format = "html", caption = "ESTIMATE Deconvolution Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
create_heatmaps(estimate_results, "ESTIMATE Results")


#TODO: check +0.001 ( it justa  shift)

####find why "Timer" wont work

# 1. Filter Low or Zero Expression Genes: Ensure genes that are not expressed (or expressed at low levels) in a significant number of samples are removed.
# 
# 2. Adjust Expression Data: Use the formula log2(TPM + 0.001) to adjust the expression data appropriately.
# 
# 3. Check for Zero Values and NAs: After adjusting the expression data, double-check that no zero values or NAs remain.
# 
# 4. Run TIMER: Once you are confident that the data is correctly prepared, you can run the TIMER analysis.

# # Load your dataset (adjust the path as needed)
# ICB_jung <- get(load(file.path(Base_dir, "data_all", "ICB_Jung__Lung__PD-(L)1.rda")))
# 
# # Extract expression data
# expr_Jung <- data.frame(assays(ICB_jung))
# expr_Jung$group <- NULL
# expr_Jung$group_name <- NULL
# expr_Jung <- as.matrix(expr_Jung)
# 
# # Step 1: Filter Low or Zero Expression Genes
# r <- as.numeric(apply(expr_Jung, 1, function(i) sum(round(i, 6) == round(log2(0.001), 6))))
# remove <- which(r > ncol(expr_Jung) * 0.5)
# expr_Jung_filtered <- expr_Jung[-remove, ]
# 
# # Step 2: Adjust Expression Data
# expr_Jung_adjusted <- log2(expr_Jung_filtered + 0.001)
# 
# # Step 3: Check for NAs
# if (any(is.na(expr_Jung_adjusted))) {
#     stop("Expression data contains NA values. Please clean your data before proceeding.")
# }
# 
# # Step 4: Check for Zero Values
# if (any(expr_Jung_adjusted == 0)) {
#     cat("Warning: Expression data still contains zero values after filtering.\n")
# }
# 
# # Run TIMER analysis
# timer_results <- tryCatch({
#     as.data.frame(deconvo_tme(eset = expr_Jung_adjusted, method = "timer"))
# }, error = function(e) {
#     cat("Error in TIMER analysis:", e$message, "\n")
#     NULL  # Return NULL if there's an error
# })
# 
# # Check results
# if (!is.null(timer_results)) {
#     cat("TIMER analysis completed successfully.\n")
#     print(head(timer_results))  # Inspect the first few rows of results
# } else {
#     cat("TIMER analysis failed. Check your input data.\n")
# }
# 


##TODO1: NEXTSTEP:
# - only scaling (by Monday)--> review
# - we have cell types ( eg.Cibersort for one dataset ,find assosiation btw cell types vs IO outcoumne( OS/PFS/Response)  )--> assosiation functions based on function ( Gneesiglogreg, os pfs)
# potentially update the function for genesigs and cell sigs.
#tehen other methos ( at least one)

##TODO2: QC for abacus, oak, poplar


##TODO3: insight- update sigs- (interestId)
##TODO4:charoenting -list of gene_listof function-
#- gensigIPSfunction ( calling two function)column #gene--

##TODO5: list of allsignitures check assisitation by the each papaer  and predicio.ca


##TODO6: github "predioR"(geneSigMean.R)- wa supdated ( fo genedsig.. wa upadted )

## TODO7:we want to add TMB score. (inspire,pola,oak, Immotion150)
# wecan generate TMB score. ICB_Common ( how we quick extract Rna fsq tmscore extract)---> check TMB score from the H4h or ICB_Common.pipline finding

## TODO8:next steps:Function piline Rna-->TMB score.

#TMB and assosiation --insight---

```

