---
title: "Deconvolution Porcess"
author: "Nasim Bondar Sahebi"
date: "2023-10-10"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## load libraries

```{r libraries}
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(immunedeconv)
library(ComplexHeatmap)
library(circlize)
library(knitr)
library(kableExtra)
library(dplyr)
library(IOBR)
#library(tools)
```

 

## Deconvolution Methods

### Overview of Dataset and Clinical Characteristics

#### Inclusion Criteria for Dataset Analysis:

- **Treatments**: Patients treated with `"PD-1/PD-L1"`, `"IO+combo"`, or `"CTLA4"` were included.
- **Cancer Types**: Only cancer types with at least 20 patients were considered. Non-solid tumors were prioritized, while solid tumors like `"Lymph_node"` were excluded.
- **Sample Types**: Focused on `"PRE"` treatment samples to assess baseline characteristics before therapy.
- **RNA-seq Data**: Only patients with RNA-seq data were included; those with microarray data were excluded.
- **Data Normalization**: RNA-seq data was normalized using `log2(TPM + 0.001)` to ensure comparable expression levels across patients.

#### Summary of Clinical Characteristics:

- **Number of Studies**: 38 (All Data), 51 (Data with Sex).
- **Number of Patients**: 2,178 (All Data), 1,758 (Data with Sex).

##### Sex Distribution:

- **Female**: 32.6% (All Data), 40.3% (Data with Sex).
- **Male**: 48.2% (All Data), 59.7% (Data with Sex).
- **Sex Not Available**: 19.3% (All Data), 0% (Data with Sex).

##### Cancer Types:

- **Lung Cancer**: 33.9% (All Data), 36.6% (Data with Sex).
- **Melanoma**: 25.4% (All Data), 23.2% (Data with Sex).
- **Bladder Cancer**: 16.9% (All Data), 16.2% (Data with Sex).
- **Other Types**: Includes smaller proportions of kidney, brain, ureteral, pancreas, and esophageal cancers.

##### Treatment Types:

- **PD-1/PD-L1**: 90.4%.
- **IO+combo**: 5.6%.
- **CTLA4**: 4%.

##### Treatment Response:

- **Non-Responders**: 55.2% (All Data), 55.6% (Data with Sex).
- **Responders**: 25.2% (All Data), 25.7% (Data with Sex).

##### Progression-Free Survival (PFS):

- **Median PFS**: 3.1 months.
- **IQR PFS**: 7.9 months (All Data), 8.6 months (Data with Sex).

##### Overall Survival (OS):

- **Median OS**: 12.2 months (All Data), 12.1 months (Data with Sex).
- **IQR OS**: 16.8 months (All Data), 16.7 months (Data with Sex).

```{r Deconvulation method for Jun Jun datastes}
# Set base directory
Base_dir <- "~/BHK lab/EcoTyper_Project/"

# Load and display Summary and Clinical datasets
kable(read.csv(file.path(Base_dir, "outputs", "Summary of All Data and Data with Sex.csv")), 
      caption = "Summary of All Data and Data with Sex") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")

kable(read.csv(file.path(Base_dir, "outputs", "Detailed_clinical_characteristics_of_the_IO_datasets.csv")), 
      caption = "Detailed Clinical Characteristics of the IO Datasets") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  scroll_box(height = "400px")

```


### 2. Generate a Table of IOBR Deconvolution Methods

Table summarizing the deconvolution methods in the  [IOBR package](https://github.com/IOBR/IOBR). Include the following columns:

1. **Method**: List the available deconvolution methods.
2. **Accepted Expression and Normalization**: Specify accepted expression types (e.g., counts, RNA-seq, TPM, etc.).
3. **PubMed ID**: Include the PubMed ID or DOI link for each method.
4. **Type of Cell**: Indicate the type of cells targeted by each method (e.g., CIBERSORT --> Immune cells).
5. **Number of Cells**: Provide the number of cell types for each method.


```{r IBOR Deconvulation methods}

methods_table <- data.frame(
  Method = c(
    "[CIBERSORT](https://cibersort.stanford.edu/)",
    "[EPIC](https://gfellerlab.shinyapps.io/EPIC_1-1/)",
    "[MCPcounter](https://github.com/ebecht/MCPcounter/blob/master/Source/License)",
    "[xCell](https://github.com/dviraran/xCell/blob/master/DESCRIPTION)",
    "[ESTIMATE](https://bioinformatics.mdanderson.org/public-software/estimate/)",
    "[TIMER](http://cistrome.org/TIMER/download.html)",
    "[quanTIseq](http://icbi.at/software/quantiseq/doc/index.html)",
    "[IPS](https://github.com/icbi-lab/Immunophenogram/blob/master/LICENSE)"
  ),
  Accepted_Expression_Normalization = c(
    "Counts, RNA-seq, TPM, FPKM",
    "RNA-seq, TPM, FPKM",
    "RNA-seq, TPM, FPKM",
    "RNA-seq, TPM, FPKM, Counts",
    "RNA-seq, FPKM",
    "RNA-seq, TPM",
    "RNA-seq, TPM",
    "RNA-seq, TPM"
  ),
  PubMed_ID = c(
    "[PMID: 25977888](https://doi.org/10.1038/nmeth.3337)",
    "[PMID: 28783075](https://doi.org/10.7554/eLife.26476)",
    "[PMID: 27798624](https://doi.org/10.1186/s13059-016-1070-5)",
    "[PMID: 29141660](https://doi.org/10.1186/s13059-017-1349-1)",
    "[PMID: 24113773](https://doi.org/10.1038/ncomms3612)",
    "[PMID: 27423106](https://doi.org/10.1186/s13059-016-1028-7)",
    "[PMID: 30992056](https://doi.org/10.1186/s13073-019-0638-6)",
    "[PMID: 28052254](https://doi.org/10.1016/j.celrep.2016.12.019)"
  ),
  Cell_Type = c(
    "Immune cells",
    "Cancer cells, Immune cells",
    "Immune cells, Stromal cells",
    "Immune cells, Stromal cells",
    "Tumor purity, Stromal and Immune cells",
    "Immune cells",
    "Tumor immune cells",
    "Immune phenotypes"
  ),
  Number_of_Cells = c(
    "22 immune cell types",
    "Multiple (B cells, T cells, Macrophages)",
    "10 cell types",
    "64 cell types",
    "Not specified",
    "6 immune cell types",
    "10 immune cell types",
    "28 immune cell types"
  ),
  License = c(
    "free for non-commercial use only",
    "free for non-commercial use only (Academic License)",
    "free [(GPL 3.0)](https://github.com/ebecht/MCPcounter/blob/master/Source/License)",
    "free [(GPL 3.0)](https://github.com/dviraran/xCell/blob/master/DESCRIPTION)",
    "free [(GPL 2.0)](https://bioinformatics.mdanderson.org/estimate/)",
    "free [(GPL 2.0)](http://cistrome.org/TIMER/download.html)",
    "free [(BSD)](https://github.com/icbi-lab/immunedeconv/blob/master/LICENSE.md)",
    "free [(BSD)](https://github.com/icbi-lab/Immunophenogram/blob/master/LICENSE)"
  ),
  Citation = c(
    "Newman, A. M., Liu, C. L., Green, M. R., Gentles, A. J., Feng, W., Xu, Y., … Alizadeh, A. A. (2015). Robust enumeration of cell subsets from tissue expression profiles. Nature Methods, 12(5), 453–457. [https://doi.org/10.1038/nmeth.3337](https://doi.org/10.1038/nmeth.3337)",
    "Racle, J., de Jonge, K., Baumgaertner, P., Speiser, D. E., & Gfeller, D. (2017). Simultaneous enumeration of cancer and immune cell types from bulk tumor gene expression data. ELife, 6, e26476. [https://doi.org/10.7554/eLife.26476](https://doi.org/10.7554/eLife.26476)",
    "Becht, E., Giraldo, N. A., Lacroix, L., Buttard, B., Elarouci, N., Petitprez, F., … de Reyniès, A. (2016). Estimating the population abundance of tissue-infiltrating immune and stromal cell populations using gene expression. Genome Biology, 17(1), 218. [https://doi.org/10.1186/s13059-016-1070-5](https://doi.org/10.1186/s13059-016-1070-5)",
    "Aran, D., Hu, Z., & Butte, A. J. (2017). xCell: digitally portraying the tissue cellular heterogeneity landscape. Genome Biology, 18(1), 220. [https://doi.org/10.1186/s13059-017-1349-1](https://doi.org/10.1186/s13059-017-1349-1)",
    "Vegesna R, Kim H, Torres-Garcia W, …, Verhaak R. (2013). Inferring tumour purity and stromal and immune cell admixture from expression data. Nature Communications 4, 2612. [http://doi.org/10.1038/ncomms3612](http://doi.org/10.1038/ncomms3612)",
    "Li, B., Severson, E., Pignon, J.-C., Zhao, H., Li, T., Novak, J., … Liu, X. S. (2016). Comprehensive analyses of tumor immunity: implications for cancer immunotherapy. Genome Biology, 17(1), 174. [https://doi.org/10.1186/s13059-016-1028-7](https://doi.org/10.1186/s13059-016-1028-7)",
    "Finotello, F., Mayer, C., Plattner, C., Laschober, G., Rieder, D., Hackl, H., …, Sopper, S. (2019). Molecular and pharmacological modulators of the tumor immune contexture revealed by deconvolution of RNA-seq data. Genome medicine, 11(1), 34. [https://doi.org/10.1186/s13073-019-0638-6](https://doi.org/10.1186/s13073-019-0638-6)",
    "Charoentong, P., et al. (2017). Pan-cancer Immunogenomic Analyses Reveal Genotype-Immunophenotype Relationships and Predictors of Response to Checkpoint Blockade. Cell Reports 18, 248-262. [https://doi.org/10.1016/j.celrep.2016.12.019](https://doi.org/10.1016/j.celrep.2016.12.019)"
  )
)

# Generate the table using kable
kable(methods_table, format = "markdown", align = "l", caption = "Deconvolution Methods in IOBR Package with License and Citation Links")

```

Here’s the corrected version of the document:

---

### CIBERSORT on **Jung Lung PD-1/PD-L1**

We begin by analyzing the **Jung Lung PD-1/PD-L1** dataset, which contains clinical characteristics of patients who received PD-1/PD-L1 treatment for lung cancer. The dataset has been normalized using RNA-seq data and includes various clinical factors such as age, sex, response to treatment, progression-free survival (PFS), and overall survival (OS).

**Important Note**: Currently, the RNA-seq data is in log2(TPM + 0.001) format. To ensure consistency, we need to convert the expression data to a positive expression range using **log2(TPM + 1)** for all samples, so all expression data is on the same scale.

We will begin the analysis with a focus on **responders (R) vs. non-responders (NR)**. The first step is to retrieve multi-assay data and analyze immune cell proportions based on the expression data.

We'll start with **CIBERSORT** to identify tumor microenvironment (TME) signatures, such as stromal or immune cell types. CIBERSORT uses a linear model to estimate the cell type distribution within the TME, which will be visualized using a heatmap.

For this analysis, we will focus on the **Jung Lung (PD-1/PD-L1)** dataset, specifically examining the TME patterns for responders (R).

**Next Steps**: We will also apply other deconvolution methods, such as xCell, EPIC, TIMER, and QUANTISEQ, though this is not a priority at this stage.


```{r IBOR Deconvolution method for Jung Jung datasets, fig.height=12, fig.width=10}

# Load the specific .rda file (ICB_Jung_Lung_PD-L1)
ICB_jung <- get(load(file.path(Base_dir, "data_all", "ICB_Jung__Lung__PD-(L)1.rda")))

# Extract expression data and convert to a matrix
expr_Jung <- data.frame(assays(ICB_jung))
expr_Jung$group <- NULL
expr_Jung$group_name <- NULL
expr_Jung <- as.matrix(expr_Jung)

# Extract clinical data
clin_jung <- data.frame(colData(ICB_jung))

# Step 1: Convert expression data from log2(TPM + 0.001) back to TPM
expr_Jung_TPM <- 2^(expr_Jung) - 0.001
expr_Jung <- log2(expr_Jung_TPM + 1)

# Run deconvolution analysis using CIBERSORT
cibersort_results <- deconvo_tme(eset = expr_Jung, method = "cibersort")
cibersort_results <- as.data.frame(cibersort_results)
rownames(cibersort_results) <- colnames(expr_Jung)
cibersort_results$ID <- NULL
colnames(cibersort_results) <- gsub("_CIBERSORT", "", colnames(cibersort_results))


kable(cibersort_results, format = "html", caption = "CIBERSORT Deconvolution Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))


# Correct column names for removal
filtered_cibersort_results <- cibersort_results %>%
select(-`P-value`, -`Correlation`, -`RMSE`)


##1. non scaled version
reorder_index <- order(clin_jung$response)
 filtered_cibersort_results <-  filtered_cibersort_results[reorder_index, ]
clin_jung <- clin_jung[reorder_index, ]

response_annotation <- HeatmapAnnotation(Response = clin_jung$response)

# Create a heatmap with scaled results
h1 <- Heatmap(t(as.matrix( filtered_cibersort_results)), name = "CIBERSORT Results (Not Scaled)", top_annotation  = response_annotation,cluster_rows = F,column_split = clin_jung$response)

## 2. scaled version
# Scale and cap the filtered CIBERSORT results between -4 and 4
scaled_cibersort_results <- as.data.frame(scale(filtered_cibersort_results))
#scaled_cibersort_results <- pmax(pmin(scaled_cibersort_results, 4), -4)

reorder_index <- order(clin_jung$response)
scaled_cibersort_results <- scaled_cibersort_results[reorder_index, ]
clin_jung <- clin_jung[reorder_index, ]

response_annotation <- HeatmapAnnotation(Response = clin_jung$response)

# Create a heatmap with scaled results
h2 <- Heatmap(t(as.matrix(scaled_cibersort_results)), name = "CIBERSORT Results (Scaled)", top_annotation  = response_annotation,cluster_rows = F,column_split = clin_jung$response)

h1
h2
# TODO: take median
```

```{r IBOR other Deconvolution  for Jung Jung datasets, fig.height=8, fig.width=10}
#TODO: you can use median of its no scaled:
common_color_scale <- colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))

# Run deconvolution analysis using epic
epic_results <- as.data.frame(deconvo_tme(eset = expr_Jung, method = "epic"))
rownames(epic_results) <- epic_results$ID 
epic_results$ID <- NULL

kable(epic_results, format = "html", caption = "Epic Deconvolution Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

reorder_index <- order(clin_jung$response)
epic_results <- epic_results[reorder_index, ]
clin_jung <- clin_jung[reorder_index, ]

epic_results_scaled <- as.data.frame(scale(epic_results))

# Create annotation for heatmap based on response
response_annotation <- HeatmapAnnotation(Response = clin_jung$response)

# Create the heatmap for scaled EPIC results
Heatmap(t(as.matrix(epic_results_scaled)), 
        name = "EPIC Results (Scaled)", 
        top_annotation = response_annotation, 
        cluster_rows = FALSE, 
        column_split = clin_jung$response, 
        col = common_color_scale)

# Run deconvolution analysis using MCPcounter
mcpcounter_results <- as.data.frame(deconvo_tme(eset = expr_Jung, method = "mcpcounter"))
rownames(mcpcounter_results) <- colnames(expr_Jung)
mcpcounter_results$ID <- NULL

kable(mcpcounter_results, format = "html", caption = "MCPcounter Deconvolution Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

mcpcounter_results <- mcpcounter_results[reorder_index, ]
mcpcounter_results_scaled <- as.data.frame(scale(mcpcounter_results))


# Create the heatmap for scaled MCPcounter results
Heatmap(t(as.matrix(mcpcounter_results_scaled)), 
        name = "MCPcounter Results (Scaled)", 
        top_annotation = response_annotation, 
        cluster_rows = FALSE, 
        column_split = clin_jung$response, 
        col = common_color_scale)

# Run deconvolution analysis using xCell
xcell_results <- as.data.frame(deconvo_tme(eset = expr_Jung, method = "xcell"))
rownames(xcell_results) <- colnames(expr_Jung)
xcell_results$ID <- NULL

kable(xcell_results, format = "html", caption = "xCell Deconvolution Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

xcell_results <- xcell_results[reorder_index, ]
xcell_results_scaled <- as.data.frame(scale(xcell_results))

# Create the heatmap for scaled xCell results
Heatmap(t(as.matrix(xcell_results_scaled)), 
        name = "xCell Results (Scaled)", 
        top_annotation = response_annotation, 
        cluster_rows = FALSE, 
        column_split = clin_jung$response, 
        col = common_color_scale)

# Run deconvolution analysis using ESTIMATE
estimate_results <- as.data.frame(deconvo_tme(eset = expr_Jung, method = "estimate"))
rownames(estimate_results) <- colnames(expr_Jung)
estimate_results$ID <- NULL

kable(estimate_results, format = "html", caption = "ESTIMATE Deconvolution Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

estimate_results <- estimate_results[reorder_index, ]
estimate_results_scaled <- as.data.frame(scale(estimate_results))

# Create the heatmap for scaled ESTIMATE results
Heatmap(t(as.matrix(estimate_results_scaled)), 
        name = "ESTIMATE Results (Scaled)", 
        top_annotation = response_annotation, 
        cluster_rows = FALSE, 
        column_split = clin_jung$response, 
        col = common_color_scale)

#TODO: check why time rnot working ( except IPS) and see what is process working , how portion across samples are estimating ( if any ML alogrithm used)
# TODO: sending for review and next step

# # Run deconvolution analysis using TIMER
# 
# timer_results <- as.data.frame(deconvo_tme(eset = expr_Jung, method = "timer"))
# rownames(timer_results) <- colnames(expr_Jung)
# 
# kable(timer_results, format = "html", caption = "TIMER Deconvolution Results") %>%
#   kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
# 
# timer_results <- timer_results[reorder_index, ]
# timer_results_scaled <- as.data.frame(scale(timer_results))
# 
# # Create the heatmap for scaled TIMER results
# Heatmap(t(as.matrix(timer_results_scaled)), 
#         name = "TIMER Results (Scaled)", 
#         top_annotation = response_annotation, 
#         cluster_rows = FALSE, 
#         column_split = clin_jung$response, 
#         col = common_color_scale)
# 
# # Run deconvolution analysis using quanTIseq
# quantiseq_results <- as.data.frame(deconvo_tme(eset = expr_Jung, method = "quantiseq"))
# rownames(quantiseq_results) <- colnames(expr_Jung)
# 
# kable(quantiseq_results, format = "html", caption = "quanTIseq Deconvolution Results") %>%
#   kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
# 
# quantiseq_results <- quantiseq_results[reorder_index, ]
# quantiseq_results_scaled <- as.data.frame(scale(quantiseq_results))
# 
# # Create the heatmap for scaled quanTIseq results
# Heatmap(t(as.matrix(quantiseq_results_scaled)), 
#         name = "quanTIseq Results (Scaled)", 
#         top_annotation = response_annotation, 
#         cluster_rows = FALSE, 
#         column_split = clin_jung$response, 
#         col = common_color_scale)


```