---
title: "Deconvolution Porcess"
author: "Nasim Bondar Sahebi"
date: "2023-11-07"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## load libraries

```{r libraries}
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(dplyr)
library(knitr)
library(kableExtra)
library(tools)
```

## Deconvolution Methods, Starting with Jung Lung PD-1/PD-L1 Dataset

Begin with one dataset, focusing on responders (R) vs. non-responders (NR).
Use a function to retrieve multi-assay data and analyze immune cell proportions based on expression data.

Start with CIBERSORT to identify tumor microenvironment (TME) signatures, such as stromal or immune cells.
The deconvolution method (CIBERSORT) will show cell type distribution based on the data.
The model used here is a linear model → visualized using a heatmap.

Also, apply other deconvolution methods like xCell, EPIC, TIMER, and QUANTISEQ (not a priority).

Focus on the Jung Lung (PD-1/PD-L1) dataset and analyze the pattern of TME for responders (R).


```{r Deconvulation method for Jun Jun datastes}

# Load the package
library(immunedeconv)
library(pheatmap)
# TODO: package "IOBR"- CRAN.

dat_icb <- get(load("~/BHK lab/Ecotyper/data/ICB_Jung__Lung__PD-(L)1.rda"))
expr_Jung <- data.frame(assays(dat_icb))
expr_Jung$group <- NULL
expr_Jung$group_name <- NULL

clin_jung <- data.frame(colData(dat_icb))

expr_Jung <- 2^expr_Jung - 0.001

# Perform deconvolution using quanTIseq
result_quanTIseq <- deconvolute(expr_Jung, method = "quantiseq")

# View the results
head(result_quanTIseq)

# Convert results to a matrix for visualization
result_matrix <- as.matrix(result_quanTIseq[, -1])  # Exclude the first column (cell type names)

# Create a heatmap
pheatmap(result_matrix, cluster_rows = TRUE, cluster_cols = TRUE, display_numbers = TRUE, 
         main = "quanTIseq Deconvolution Heatmap")


# TODO: Positive expression. For all expression data, ensure it is on the same scale using log2(TPM + 1).
# TODO: Create a table specifying the normalized expression that is accepted:
# 1. Columns (e.g., counts, RNA-seq, TPM, etc.)
# 2. Generate Table:
#      - Method: Find available deconvolution methods in the "IOBR" package and include the following:
#      - Accepted expression and normalization (e.g., counts, RNA-seq, TPM, etc.)
#      - PubMed ID for each method
#      - Type of cell (e.g., CIBERSORT --> Immune cells, other combinations)
#      - Number of cells for each method.


# TODO: Perfrom each method+ Create a heatmap for each deconvolution method for a dataset.
# TODO: Write a separate script for comparing deconvolution methods against data.

```